Standartox: Standardizing toxicity data

### 2020-05-04 article/article.tex

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{article/figures/results_variability.png}
    \caption{Violin plots of of test results (EC\textsubscript{50}) in Standartox illustrating (\textbf{A}) differential variability and data distribution between species (i.e. \textit{Xenopus laevis} - Amphibia, \textit{Raphidocelis subcapitata} - Algae, \textit{Oncorhynchus mykiss} - fish, \textit{Lemna minor} - invertebrate) for the chemical atrazine in 96 h tests, (\textbf{B}) how the variability in toxicity tests with zinc sulfate and \textit{Daphnia magna} varies with test duration and (\textbf{C}) high variability that is not explained by the available test characteristics in the case of cupric sulfate tested on \textit{Pimephales promelas} for 96 h. Red dots depict Standartox geometric mean estimates and black dots show raw data for each group. To facilitate readability, data points are randomly scattered along a hypothetical y-axis and are greyed out if within the violins.}
    \label{fig:stx-variability}
\end{figure}

@@

\subsection{Processing}
Standartox downloads the quarterly released ECOTOX database and builds it into a local PostgreSQL database. Subsequently, Structured Query Language (SQL) functions for further processing the data are implemented. In addition lookup tables that enable the conversion of units such as duration and concentration are created. A meta-table providing information such as the release version of the ECOTOX database is added. Then, provided Chemical Abstracts Service Numbers (CAS) and taxonomic names are used to query additional information from publicly available databases on chemicals and organisms, respectively. This includes the Compendium of Pesticide Common Names \citep{wood_compendium_2019}, the Chemical Entities of Biological Interest (ChEBI) database \citep{hastings_chebi_2016}, the Chemical Identifier Resolver (CIR) service \citep{nationalinstitutesofhealthnih_chemical_2019}, the Pubchem database \citep{kim_pubchem_2016}, Eurostat \citep{europeancommission_eurostat_2019} and Wikidata \citep{vrandecic_wikidata_2014} for chemicals and the World Register of Marine Species (WoRMS) \citep{wormseditorialboard_world_2020} and the Global Biodiversity Information Facility (GBIF) \citep{gbiftheglobalbiodiversityinformationfacility_what_2020} (Table~\ref{tab:data-base-additional}) for habitat and spatial distribution of organisms. Given that taxonomic names can be ambiguous, e.g. the genus \textit{Eisenia} can refer to an algae and a worm, we first match the taxa names against specific database identifiers (e.g. WoRMS \citep{wormseditorialboard_world_2020}) and subsequently check their accordance with the underlying ECOTOX data taxonomy. Then we query the actual data by using the identifiers. In a next step the data is added to Standartox to enable filtering for specific chemical roles (e.g. drug, metal, pesticide, personal care product) and classes (e.g. pyrethroid, carbamate) as well as spatial distribution (i.e. continents) and habitat preferences (e.g. freshwater) of individual taxa. Taxa that were not identified to at least genus level are excluded, because relative toxicity comparisons have been shown to be not meaningful for higher taxonomic levels \citep{rainbow_trace_2002, buchwalter_differences_2005, malaj_physiological_2012}. Finally, the Standartox data set is compiled, which includes the harmonisation of data, e.g. through conversion of test concentration and duration units. 1237 distinct concentration units are converted to six harmonized ones (i.e. g/l, g/m2, ppb, g/g, l/l, l/m2) when conversion is possible. Likewise, the 126 distinct duration units are converted to hours whenever this is unambiguously possible. To guarantee appropriate unit conversion and harmonisation, we compared the results of an automated unit conversion to a manual one for each of the distinct concentration and duration units. This assures that 652 of the 1237 units  (95.3 \% of the data) are converted correctly. The remainder could not be converted.  Furthermore, the units are cleaned, for example through removing additional information in the field such as \textit{food}, \textit{soil}, \textit{ai} that are also coded in other variables and hinder the processing of units. Concentrations that are given as rates such as per day (e.g. mg/kg/day) are multiplied by the days of the test and then converted. Experimental endpoints are restricted to three groups, namely NOEX, LOEX and XX\textsubscript{50}. Other endpoints, such as Bioconcentration factors, non-half maximal effective concentrations (e.g. IC\textsubscript{10}, EC\textsubscript{25}, LD\textsubscript{99}) or maximum acceptable toxicant concentrations are removed. Along with that, a catalog, listing all distinct entries and value ranges, for categorical and continuous variables, respectively, is created. Finally, we run quality control scripts that check the accuracy of the data.

CONTINUE HERE

The compiled Standartox data set together with the catalog is exported and accessible via the web application and the API, through the R package.

@@

\subsection{Application methods}
When accessed, the web application and the API load the compressed serialized Standartox data into memory and allow the user to interact with it. The user can then call the functions \textit{stx\_filter()} and \textit{stx\_aggregate()} that filter and aggregate the data according to specific parameters (Table~\ref{tab:app-parameters}). The interactive web application is built in R using the shiny framework, which runs with the help of a shiny server \citep{R-shiny} on \url{standartox.uni-landau.de}. The API is built by using the R package \textit{plumber} \citep{R-plumber}, which allows for the creation of Representational State Transfer (REST) APIs from R. REST is a software architectural style that defines web service communication rules. The API is reachable via the URL \url{http://139.14.20.252} and port 8000. Four API-endpoints (\textit{/catalog}, \textit{/filter}, \textit{/aggregate} and \textit{/meta}) can be queried (Table~\ref{tab:api-endpoints}). The \textit{/catalog} API-endpoint returns a JavaScript Object Notation (JSON) file containing a catalog of possible filter parameters to choose from. The \textit{/filter} returns the filtered Standartox table as a compressed serialized binary file created by the R \textit{fst} package \citep{R-fst}, to reduce size and allow for fast user queries. The \textit{/aggregate} API-endpoint returns an R function allowing the client to aggregate the filtered data locally. Lastly, the \textit{/meta} API-endpoint returns a JSON file with meta information, such as the timestamp of the request and the used Standartox version. The API is designed to be used with the R-package \textit{standartox} and therefore uses serialization methods specific to R (rds() from the R package \textit{base} and fst() from the package \textit{fst}). To facilitate the API usage the R-package \textit{standartox} is created.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{yes} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendix
\section{}
\begin{table}[H]
    \caption{Table of additionally queried publicly available databases and their URLs.}
    \label{tab:data-base-additional}
    \centering
\begin{tabular}{cc}
    \toprule
    \textbf{Database} & \textbf{URL} \\ 
    \midrule
    \makecell{Chemical Entities of \\ Biological Interest (ChEBI)} & \url{https://www.ebi.ac.uk/chebi} \\
    Chemical Identifier Resolver & \url{https://cactus.nci.nih.gov/chemical/structure} \\[0.5cm]
    ChemSpider & \url{http://www.chemspider.com}    \\[0.5cm]
    Eurostat & \url{https://ec.europa.eu/eurostat/home} \\[0.5cm]
    PubChem & \url{https://pubchem.ncbi.nlm.nih.gov} \\[0.5cm]
    Wikidata & \url{https://www.wikidata.org/wiki/Wikidata:Main_Page} \\[0.5cm]
    \makecell{Global Biodiversity \\ Information Facility (GBIF)} & \url{https://www.gbif.org} \\[0.5cm]
    \makecell{World Register of \\ Marine Species (WoRMS)} & \url{http://marinespecies.org} \\
    \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[H]
    \caption{Application programming interface (API) endpoints, HTTP methods, Requests and Response objects. JSON - Javascript opject notation file}
    \label{tab:api-endpoints}
    \centering
\begin{tabular}{cccc}
    \toprule
    \textbf{Endpoint} & \textbf{HTTP Method} & \textbf{Request} & \textbf{Response} \\
    \midrule
    /catalog & POST & Standartox version string & Catalog object (JSON) \\
    /filter & POST & Standartox filter parameters & Filtered Standartox data (serialized) \\
    /aggregate & GET & / & Aggregate function (serialized) \\
    /meta & POST & Standartox version string & Meta data on request (JSON) \\
    \bottomrule
\end{tabular}
\end{table}

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{article/figures/results_variability.png}
    \caption{Violin plots of of test results (EC\textsubscript{50}) in Standartox illustrating (\textbf{A}) differential variability and data distribution between species (i.e. \textit{Xenopus laevis} - Amphibia, \textit{Raphidocelis subcapitata} - Algae, \textit{Oncorhynchus mykiss} - fish, \textit{Lemna minor} - invertebrate) for the chemical atrazine in 96 h tests, (\textbf{B}) how the variability in toxicity tests with zinc sulfate and \textit{Daphnia magna} varies with test duration and (\textbf{C}) high variability that is not explained by the available test characteristics in the case of cupric sulfate tested on \textit{Pimephales promelas} for 96 h. Red dots depict Standartox geometric mean estimates and black dots show raw data for each group. Red error bars show the standard deviation of the geometric mean. To facilitate readability, data points are randomly scattered along a hypothetical y-axis and are greyed out if within the violins.}
    \label{fig:stx-variability}
\end{figure}

@@

\subsection{Processing}
Standartox downloads the quarterly released ECOTOX database and builds it into a local PostgreSQL database. Subsequently, Structured Query Language (SQL) functions for further processing the data are implemented. In addition lookup tables that enable the conversion of units such as duration and concentration are created. A meta-table providing information such as the release version of the ECOTOX database is added. Then, provided Chemical Abstracts Service Numbers (CAS) and taxonomic names are used to query additional information from publicly available databases on chemicals and organisms, respectively. This includes the Compendium of Pesticide Common Names \citep{wood_compendium_2019}, the Chemical Entities of Biological Interest (ChEBI) database \citep{hastings_chebi_2016}, the Chemical Identifier Resolver (CIR) service \citep{nationalinstitutesofhealthnih_chemical_2019}, the Pubchem database \citep{kim_pubchem_2016}, Eurostat \citep{europeancommission_eurostat_2019} and Wikidata \citep{vrandecic_wikidata_2014} for chemicals and the World Register of Marine Species (WoRMS) \citep{wormseditorialboard_world_2020}, the Global Biodiversity Information Facility (GBIF) \citep{gbiftheglobalbiodiversityinformationfacility_what_2020} and the freshwaterecology.info database \citep{schmidt-kloiber_www_2015} (Table~\ref{tab:data-base-additional}) for habitat and spatial distribution of organisms. Given that taxonomic names can be ambiguous, e.g. the genus \textit{Eisenia} can refer to an algae and a worm, we first match the taxa names against specific database identifiers (e.g. WoRMS \citep{wormseditorialboard_world_2020}) and subsequently check their accordance with the underlying ECOTOX data taxonomy. Then we query the actual data by using the identifiers. In a next step the data is added to Standartox to enable filtering for specific chemical roles (e.g. drug, metal, pesticide, personal care product) and classes (e.g. pyrethroid, carbamate) as well as spatial distribution (i.e. continents) and habitat preferences (e.g. freshwater) of individual taxa. Taxa that were not identified to at least genus level are excluded, because relative toxicity comparisons have been shown to be not meaningful for higher taxonomic levels \citep{rainbow_trace_2002, buchwalter_differences_2005, malaj_physiological_2012}. Finally, the Standartox data set is compiled, which includes the harmonisation of data, e.g. through conversion of test concentration and duration units. 1237 distinct concentration units are converted to six harmonized ones (i.e. g/l, g/m2, ppb, g/g, l/l, l/m2) when conversion is possible. Likewise, the 126 distinct duration units are converted to hours whenever this is unambiguously possible. To guarantee appropriate unit conversion and harmonisation, we compared the results of an automated unit conversion to a manual one for each of the distinct concentration and duration units. This assures that 652 of the 1237 units (95.3 \% of the data) are converted correctly. The remainder could not be converted and is removed.  Furthermore, the units are cleaned, for example through removing additional information in the field such as \textit{food}, \textit{soil}, \textit{ai} that are also coded in other variables and hinder the processing of units. Concentrations that are given as rates such as per day (e.g. mg/kg/day) are multiplied by the days of the test and then converted. Experimental endpoints are restricted to three groups, namely NOEX, LOEX and XX\textsubscript{50}. Other endpoints, such as Bioconcentration factors, non-half maximal effective concentrations (e.g. IC\textsubscript{10}, EC\textsubscript{25}, LD\textsubscript{99}) or maximum acceptable toxicant concentrations are removed. Along with that, a catalog, listing all distinct entries and value ranges, for categorical and continuous variables, respectively, is created. The compiled Standartox data set together with the catalog is exported and accessible via the web application and the API, through the R package.

@@

\subsection{Application methods}
When accessed, the web application and the API load the compressed serialized Standartox data into memory and allow the user to interact with it. The user can then call the functions \textit{stx\_filter()} and \textit{stx\_aggregate()} that filter and aggregate the data according to specific parameters (Table~\ref{tab:app-parameters}). The interactive web application is built in R using the shiny framework, which runs with the help of a shiny server \citep{R-shiny} on \url{standartox.uni-landau.de}. The API is built by using the R package \textit{plumber} \citep{R-plumber}, which allows for the creation of Representational State Transfer (REST) APIs from R. REST is a software architectural style that defines web service communication rules. The API is reachable via the URL \url{http://139.14.20.252} and port 8000. Three API-endpoints (\textit{/catalog}, \textit{/filter}, and \textit{/meta}) can be queried (Table~\ref{tab:api-endpoints}). The \textit{/catalog} API-endpoint returns a JavaScript Object Notation (JSON) file containing a catalog of possible filter parameters to choose from. The \textit{/filter} returns the filtered Standartox table as a compressed serialized binary file created by the R \textit{fst} package \citep{R-fst}, to reduce size and allow for fast user queries. Lastly, the \textit{/meta} API-endpoint returns a JSON file with meta information, such as the timestamp of the request and the used Standartox version. The API is designed to be used with the R-package \textit{standartox} and therefore uses serialization methods specific to R (rds() from the R package \textit{base} and fst() from the package \textit{fst}). To facilitate the API usage the R-package \textit{standartox} is created.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{yes} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendix
\section{}
\begin{table}[H]
    \caption{Table of additionally queried publicly available databases and their URLs.}
    \label{tab:data-base-additional}
    \centering
\begin{tabular}{cc}
    \toprule
    \textbf{Database} & \textbf{URL} \\ 
    \midrule
    \makecell{Chemical Entities of \\ Biological Interest (ChEBI)} & \url{https://www.ebi.ac.uk/chebi} \\
    Chemical Identifier Resolver & \url{https://cactus.nci.nih.gov/chemical/structure} \\[0.5cm]
    ChemSpider & \url{http://www.chemspider.com}    \\[0.5cm]
    Eurostat & \url{https://ec.europa.eu/eurostat/home} \\[0.5cm]
    PubChem & \url{https://pubchem.ncbi.nlm.nih.gov} \\[0.5cm]
    Wikidata & \url{https://www.wikidata.org/wiki/Wikidata:Main_Page} \\[0.5cm]
    \makecell{Global Biodiversity \\ Information Facility (GBIF)} & \url{https://www.gbif.org} \\[0.5cm]
    \makecell{World Register of \\ Marine Species (WoRMS)} & \url{http://marinespecies.org} \\
    freshwaterecology.info & \url{www.freshwaterecology.info} \\
    \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[H]
    \caption{Application programming interface (API) endpoints, HTTP methods, Requests and Response objects. JSON - Javascript opject notation file}
    \label{tab:api-endpoints}
    \centering
\begin{tabular}{cccc}
    \toprule
    \textbf{Endpoint} & \textbf{HTTP Method} & \textbf{Request} & \textbf{Response} \\
    \midrule
    /catalog & POST & Standartox version string & Catalog object (JSON) \\
    /filter & POST & Standartox filter parameters & Filtered Standartox data (serialized) \\
    /meta & POST & Standartox version string & Meta data on request (JSON) \\
    \bottomrule
\end{tabular}
\end{table}
